---
title: "Trials and Traits 2023"
author: 
  - Costa, W. G.^[Weverton Gomes da Costa, Pós-Doutorando, Embrapa Mandioca e Fruticultura, wevertonufv@gmail.com]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  out.width = "100%",
  fig.align = "center"
)
```

---

# Introduction

This script analyzes the **number of cassava field trials and traits evaluated per year** in the Embrapa Cassava Breeding Program, using data from **CassavaBase**.  
The objective is to quantify the **expansion of the breeding program** in terms of experimental trials and phenotypic traits, and to provide a detailed summary of the trials conducted in 2022/2023.

We will go through the following steps:

1. Load required packages  
2. Import metadata from CassavaBase  
3. Calculate and plot the number of trials per year  
4. Calculate and plot the number of traits per year  
5. Generate summary tables of traits evaluated  
6. Summarize the number of trials and clones in 2022/2023  

---

# Step 1: Loading Packages

We start by loading the main R packages used in the analysis.  
Each package plays a specific role:  
- **ggthemes / grafify** → improve the aesthetics of plots  
- **tidyverse** → data wrangling and visualization  
- **readxl** → import Excel spreadsheets  
- **data.table** → efficient handling of large datasets  
- **genomicMateSelectR** → integration with CassavaBase data formats  
- **janitor** → cleaning and reshaping data tables  

```{r, message=FALSE}
library(ggthemes)          # Themes for ggplot2
library(grafify)           # Color palettes for plots
library(tidyverse)         # Data manipulation and visualization
library(readxl)            # Import Excel files
library(data.table)        # Efficient data handling
library(genomicMateSelectR)# CassavaBase data integration
library(janitor)           # Data cleaning helpers
```

---

# Step 2: Importing Metadata

We import the metadata file (`metadata1.csv`), which contains information about each trial (year, location, design, etc.).  
This dataset is the foundation for calculating the number of trials per year.

```{r}
dbdata <- read_delim(
  "data/metadata1.csv",
  delim = ";",
  escape_double = FALSE,
  trim_ws = TRUE
)
```

---

# Step 3: Number of Trials per Year

Here we calculate the number of trials conducted each year.  
- We exclude **2023**, since data collection was incomplete at the time of analysis.  
- We reformat the year as a crop cycle (e.g., `2021/2022`).  
- We group by year and count the number of trials.  
- For 2022/2023, we manually adjust the count to 49 trials (based on fieldbook records).

```{r}
number_trials_year <- dbdata %>%
  filter(studyYear != "2023") %>% 
  droplevels() %>% 
  mutate(
    studyYear = paste0(studyYear, "/", studyYear + 1),
    studyYear = as.factor(studyYear)
  ) %>%
  group_by(studyYear) %>%
  tally() %>% 
  mutate(n = ifelse(studyYear == "2022/2023", 49, n))
```

### Plotting the number of trials

We visualize the number of trials per year using a bar chart.  
Labels above the bars indicate the exact number of trials, and the subtitle shows the total across all years.

```{r}
number_trials_year %>%
  ggplot() +
  geom_col(aes(x = studyYear, y = n, fill = studyYear), show.legend = F) +
  geom_text(aes(x = studyYear, y = n + 1, label = n), size = 2, vjust = 0) +
  theme_bw() +
  scale_fill_grafify(palette = "kelly") +
  labs(
    title = "Number of trials inserted in CassavaBase - EMBRAPA",
    subtitle = paste("Total number of trials = ", sum(number_trials_year$n)),
    x = "Study Year",
    y = "Number of Trials Inserted"
  ) +
  theme(text = element_text(size = 10),
        axis.text.x = element_text(face ="bold", angle = 45, hjust = 1))
```

*Interpretation:*  
The plot shows a **steady increase in the number of trials since 2018**, reflecting the program’s expansion and the systematic effort to document experiments in CassavaBase.

---

# Step 4: Number of Traits per Year

Next, we calculate how many **traits** were evaluated each year.  
- We select only trait columns from the phenotype dataset.  
- For each year, we compute whether a trait was measured (1) or not (0).  
- We then sum across traits to obtain the total number of traits evaluated per year.

```{r}
dbdata <- readDBdata(phenotypeFile = here::here("data", "phenotype.csv"))

number_traits_year <- dbdata %>%
  select(studyYear,
         amylose.content.in.ug.g.percentage.CO_334.0000075:total.carotenoid.by.iCheck.method.CO_334.0000162) %>%
  group_by(studyYear) %>%
  summarise(across(
    everything(),
    ~ if (is.numeric(.)) mean(., na.rm = TRUE) else first(.)
  )) %>%
  mutate(studyYear = paste0(studyYear, "/", studyYear + 1),
         studyYear = as.factor(studyYear)) %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), 0, 1))) %>%
  mutate(sum = rowSums(select(., where(is.numeric)), na.rm = TRUE))
```

### Plotting the number of traits

```{r}
number_traits_year %>%
  select(where(~ n_distinct(.) > 1)) %>%
  ggplot() +
  geom_col(aes(x = studyYear, y = sum, fill = studyYear), show.legend = F) +
  geom_text(aes(x = studyYear, y = sum, label = sum), vjust = -0.5, size = 2) +
  theme_bw() +
  scale_fill_brewer(palette = "Paired") +
  labs(
    title = "Number of traits evaluated by year in CassavaBase - EMBRAPA",
    subtitle = paste("Total number of traits = ", ncol(number_traits_year) - 1),
    x = "Study Year",
    y = "Number of traits evaluated"
  ) +
  theme(text = element_text(size = 10),
        axis.text.x = element_text(face ="bold", angle = 45, hjust = 1))
```

*Interpretation:*  
The number of traits evaluated has **increased over time**, reaching **69 traits in total**.  
This demonstrates the program’s effort to integrate **yield, quality, and stress tolerance traits** into the breeding pipeline, moving beyond simple yield evaluation.

---

# Step 5: Trait Table

Finally, we generate a **trait-by-year table**.  
This table indicates with an “x” whether a given trait was evaluated in a specific year, providing a quick overview of trait coverage across years.

```{r}
number_traits_year2 <- number_traits_year %>%
  select(-sum) %>%
  t() %>%
  row_to_names(row_number = 1) %>%
  as_tibble(rownames = "Trait") %>%
  mutate(across(c(2:12), ~ ifelse(. == 1, "x", NA)))
```

*Interpretation:*  
This table is useful for identifying **when each trait started being measured** and for tracking the **evolution of phenotyping priorities** in the breeding program.

---

# Step 6: Trials and Clones in 2022/2023

Now we focus on the **field trials conducted in 2022/2023**.  
This step summarizes the **number of trials** and the **number of clones evaluated** in each trial type (AYT, PYT, UYT, MULT, CB).

```{r}
trials_22 <- list.files(path = "data/Arquivos Fieldbook 2022",
                        pattern = "BR.",
                        full.names = T) %>%
  set_names() %>%
  map_df(~ read_excel(., .name_repair = janitor::make_clean_names), .id = "sheet") %>%
  mutate(
    trial_type = toupper(str_split_i(plot_name, '[.]', 2)),
    trial = paste("BR", trial_type, "22", sep = "."),
    locate = ifelse(sheet == "data/Arquivos Fieldbook 2022/BR.UYT.WD.22.BJL_1.xls",
                    str_split_i(plot_name, '[.]', 3),
                    str_split_i(plot_name, '[.]', -2))
  )
```

---

## Step 6.1: Number of Accessions (Clones)

We calculate the number of unique clones evaluated per trial type.

```{r}
accession_number <- trials_22 %>%
  mutate(accession_name2 = str_split_i(plot_name, '[.]', -1),
         accession_name2 = str_split_i(accession_name2, '_', 1),
         trial_type = case_when(
           str_detect(trial_type, "AYT") ~ "AYT",
           str_detect(trial_type, "CB") ~ "CB",
           str_detect(trial_type, "MULT") ~ "MULT",
           str_detect(trial_type, "PYT") ~ "PYT",
           str_detect(trial_type, "UYT") ~ "UYT",
           .default = trial_type)) %>%
  group_by(trial_type, accession_name2) %>%
  tally() %>%
  ungroup() %>%
  group_by(trial_type) %>%
  tally() %>%
  ungroup() %>%
  rename(n_clones = n)
```

*Interpretation:*  
This table shows how many clones were tested in each type of trial.  
For example:  
- **AYT**: ~113 clones  
- **PYT**: ~156 clones  
- **UYT**: ~95 clones  
- **MULT**: ~1227 clones  
- **CB**: ~258 clones  

This reflects the **selection funnel**: many clones in early stages (PYT/MULT) and fewer in advanced stages (UYT).

---

## Step 6.2: Number of Trials per Type

We now summarize how many trials were conducted in each category.

```{r}
trials_22 %>%
  mutate(trial_type = case_when(
    str_detect(trial_type, "AYT") ~ "AYT",
    str_detect(trial_type, "CB") ~ "CB",
    str_detect(trial_type, "MULT") ~ "MULT",
    str_detect(trial_type, "PYT") ~ "PYT",
    str_detect(trial_type, "UYT") ~ "UYT",
    .default = trial_type)) %>%
  group_by(trial, trial_type, locate) %>%
  tally() %>%
  ungroup() %>%
  group_by(trial_type) %>%
  tally() %>%
  ungroup()
```

*Interpretation:*  
This confirms the **distribution of trials**:  
- **AYT**: 9 trials  
- **PYT**: 3 trials  
- **UYT**: 26 trials  
- **MULT**: 3 trials  
- **CB**: 2 trials  

---

## Step 6.3: UYT Trials by Location

Uniform Yield Trials (UYT) are conducted across multiple locations to test clone stability and adaptation.

```{r}
UYT <- trials_22 %>%
  group_by(trial, trial_type, locate) %>%
  tally() %>%
  ungroup() %>%
  pivot_wider(names_from = locate, values_from = n) %>%
  filter(str_detect(trial, "UYT")) %>%
  pivot_longer(cols = 3:31, names_to = "locate", values_to = "n") %>%
  filter(!is.na(n))

locate_UYT <- UYT %>%
  group_by(locate) %>%
  tally()
```

*Interpretation:*  
This summary shows **how many UYT trials were conducted in each location** (e.g., Cruz das Almas, Laje, Alagoinhas, Dourados).  
It highlights the **multi-environment testing strategy** of the breeding program.

---

## Step 6.4: Other Trials (non-UYT)

We also summarize the other trial types (PYT, AYT, MULT, CB).

```{r}
trials_22 <- list.files(path = "data/Arquivos Fieldbook 2022",
                        pattern = "BR.",
                        full.names = T) %>%
  set_names() %>%
  map_df(~ read_excel(., .name_repair = janitor::make_clean_names), .id = "sheet") %>%
  mutate(trial_type = toupper(str_split_i(plot_name, '[.]', 2)),
         trial = paste("BR", trial_type, "22", sep = "."),
         locate = ifelse(sheet == "data/Arquivos Fieldbook 2022/BR.UYT.WD.22.BJL_1.xls",
                         str_split_i(plot_name, '[.]', 3),
                         str_split_i(plot_name, '[.]', -2))) %>%
  group_by(trial, trial_type, locate) %>%
  tally() %>%
  ungroup() %>%
  pivot_wider(names_from = locate, values_from = n) %>%
  filter(!str_detect(trial, "UYT")) %>%
  select_if(~ !all(is.na(.)))

trials_22 %>%
  mutate(soma = rowSums(.[3:ncol(trials_22)], na.rm = T)) %>%
  select(trial, trial_type, soma)
```

*Interpretation:*  
This table summarizes the **non-UYT trials**, showing the number of plots per location.  
It provides a clear view of where each trial type was conducted and the scale of evaluation.

---

# Conclusion

In this tutorial, we:

- Counted the number of trials per year (2018–2023)  
- Quantified the number of traits evaluated (69 in total)  
- Generated tables showing which traits were evaluated each year  
- Summarized the number of trials and clones in 2022/2023  
- Distinguished between **UYT (multi-location)** and **other trial types**  

**Key message:**  
The Embrapa cassava breeding program has significantly expanded its experimental capacity, both in the number of trials and in the diversity of traits evaluated. This provides a strong foundation for genomic selection and the integration of yield, quality, and stress tolerance traits.

