---
title: "Phenotype Data"
author: 
  - Costa, W. G.^[Weverton Gomes da Costa, Pós-Doutorando, Embrapa Mandioca e Fruticultura, wevertonufv@gmail.com]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

```{r}
library(gt)
library(tidyverse)
library(readxl)
library(data.table)
library(genomicMateSelectR)
```

Created C0 population

```{r}
exclud_c0 <-
  c(
    "BGM-0830-TVER",
    "BGM-0611",
    "BGM-06111",
    "BGM-0022 T.Bco",
    "BGM-0022 T.Rx",
    "BGM-0113 RcxCr",
    "BGM-0113 RcxRos",
    "BGM-0115 TB",
    "BGM-0115 TR",
    "BGM-0150A",
    "BGM-0152A",
    "BGM-0152-B",
    "BGM-0780(B)",
    "BGM-0780(E)",
    "BGM-0830",
    "BGM-0830 TB",
    "BGM-0856 CCr",
    "BGM-0856 CRx",
    "BGM-1176(M)",
    "BGM-1262C",
    "BGM-1262E",
    "BGM-1502(M)",
    "BGM-1722(D)",
    "BGM-1759(B)",
    "BGM-0150-B",
    "BGM-0082",
    "BGM-0165",
    "BGM-0177",
    "Manzhitil",
    "BGM-0271",
    "BGM-0276",
    "BGM-0312",
    "BGM-0365",
    "BGM-0890",
    "BGM-0927",
    "BGM-0931",
    "BGM-1186",
    "BGM-1212",
    "BGM-1218",
    "BGM-1490",
    "BGM-0053",
    "BGM-0307",
    "BGM-0669",
    "BGM-1376",
    "BGM-0082?",
    "BGM-0271?",
    "BGM-0276?",
    "BGM-0290?",
    "BGM-0365?",
    "BGM-0655.1?",
    "BGM-0737?",
    "BGM-0869?",
    "BGM-0895?",
    "BGM-1028?",
    "BGM-1212?",
    "BGM-0043?",
    "BGM-0144?",
    "BGM-0165.1?",
    "BGM-0165.2?",
    "BGM-0182?",
    "BGM-0187?",
    "BGM-0533?",
    "BGM-0713?",
    "BGM-0890?",
    "BGM-0931?",
    "BGM-1218?",
    "BGM-1359?",
    "Manzhitil?",
    "SE-9",
    "SE-1",
    "SE-2",
    "SE-3",
    "SE-8",
    "BGM-0655.2?",
    "BGM-0077?",
    "BGM-0929?",
    "BGM-1025?",
    "BGM-1360?",
    "BGM-1511?",
    "BLANK__G12"
  )

`%!in%` <- negate(`%in%`)

C0 <- read_delim(
  "data/TP-BR-18-EMBRAPA-Nextgen.csv",
  delim = ";",
  escape_double = FALSE,
  trim_ws = TRUE
) %>%
  filter(tissue_id %!in% exclud_c0)
```

Data included phenotype and metadata from cassava base

```{r}
dbdata <-
  readDBdata(
    phenotypeFile = here::here("data", "phenotype (1).csv"),
    metadataFile = here::here("data", "metadata (1).csv")
  ) %>%
  dplyr::rename(
    FSW = fresh.shoot.weight.measurement.in.kg.per.plot.CO_334.0000016,
    FRW = fresh.storage.root.weight.per.plot.CO_334.0000012,
    DMCg = dry.matter.content.by.specific.gravity.method.CO_334.0000160,
    DMCo = dry.matter.content.percentage.CO_334.0000092,
    NOHAV = plant.stands.harvested.counting.CO_334.0000010,
    StC = starch.content.percentage.CO_334.0000071,
    PA = plant.architecture.visual.rating.1.5.CO_334.0000099,
    DRY = dry.yield.CO_334.0000014
  ) %>%
  mutate(
    Pop = case_when(
      germplasmName %like% "BR-20GS-" |
        germplasmSynonyms %like% "BR-20GS-" |
        germplasmName %like% "BR-19GS-" |
        germplasmSynonyms %like% "BR-19GS-" ~ "C2",
      germplasmName %like% "BR-18GS-" |
        germplasmSynonyms %like% "BR-18GS-" ~ "C1",
      germplasmName %in% C0$tissue_id |
        germplasmSynonyms %in% C0$tissue_id ~ "C0"
    )
  ) %>%
  mutate_if(is.character, as.factor) %>% 
  select(-numberBlocks, -numberReps)
```


```{r}
clone_pop <- dbdata %>% 
  group_by(Pop, germplasmName) %>% 
  tally() %>% 
  select(Pop, germplasmName)

clone_pop %>% 
  group_by(Pop) %>% 
  tally()

clone_pop %>% 
  group_by(germplasmName) %>% 
  tally()

write.table(clone_pop, "clone_pop.csv", sep=",", row.names = F)
```


## Count numbers of Blocks, Reps, check	and test

```{r}
numberBlocks <- dbdata %>% 
  group_by(studyName, blockNumber) %>% 
  tally() %>% 
  tally() %>% 
  dplyr::rename(numberBlocks = n)

numberReps <- dbdata %>% 
  group_by(studyName, replicate) %>% 
  tally() %>% 
  tally() %>% 
  dplyr::rename(numberReps = n) 
  
dbdata <- dbdata %>% 
  full_join(numberBlocks) %>% 
  full_join(numberReps)
```

```{r}
dbdata %>% dplyr::count(observationLevel)
dbdata %>%
  dplyr::count(studyName,
               trialType,
               studyDesign,
               numberBlocks,
               numberReps,
               entryType) %>%
  spread(entryType, n) %>%
  gt()  %>%
  tab_options(table.font.size = pct(75))
```


Filter BR.PTBAGRec.19.Candeal, there are only 1 rep and 1 block.

```{r}
dbdata	<- dbdata %>% 
  filter(studyName != "BR.PTBAGRec.19.Candeal") %>% 
  droplevels()
```

## check if the replicate and blockNumber columns reliably distinguish complete and incomplete blocks in the data.

```{r}
dbdata %>%
  group_by(studyName) %>%
  dplyr::summarize(N_replicate = length(unique(replicate)),
                   N_blockNumber = length(unique(blockNumber))) %>%
  gt() %>% tab_options(table.font.size = pct(75))
```

## The question is, are complete replications of the experiment indicated by replicate and incomplete sub-blocks represented by blockNumber

```{r}
dbdata %>%
  group_by(studyName) %>%
  dplyr::summarize(
    N_replicate = length(unique(replicate)),
    N_blockNumber = length(unique(blockNumber)),
    doRepsEqualBlocks = all(replicate == blockNumber)
  ) %>% 
  gt() %>% tab_options(table.font.size = pct(75))
```

## to check that the replicate column definitely means complete blocks. If the numbers are very similar for all trials, indicating complete blocks.

```{r}
# the overall number of accessions per trial
dbdata %>%
  group_by(studyName) %>%
  dplyr::summarize(N_accession = length(unique(germplasmName))) %>%
  # the average number of accessions per replicate per trial
  left_join(
    dbdata %>%
      group_by(studyName, replicate) %>%
      dplyr::summarize(N_accession = length(unique(germplasmName))) %>%
      group_by(studyName) %>%
      dplyr::summarize(avgAccessionsPerReplicate = ceiling(mean(N_accession)))
  ) %>%
  gt() %>% tab_options(table.font.size = pct(75))
```

## One more: look at the min, mean and max number of accessions per blockNumber

```{r}
# the overall number of accessions per trial
dbdata %>%
  group_by(studyName) %>%
  dplyr::summarize(N_accession = length(unique(germplasmName))) %>%
  left_join(
    dbdata %>%
      group_by(studyName, replicate, blockNumber) %>%
      dplyr::summarize(N_accession = length(unique(germplasmName))) %>% ungroup() %>%
      group_by(studyName) %>%
      dplyr::summarize(
        minAccessionsPerBlock = ceiling(min(N_accession)),
        avgAccessionsPerBlock = ceiling(mean(N_accession)),
        maxAccessionsPerBlock = ceiling(max(N_accession))
      )
  ) %>%
  gt() %>% tab_options(table.font.size = pct(75))
```

## to create explicitly nested design variables (yearInLoc, trialInLocYr, repInTrial, blockInTrial).

```{r}
dbdata %<>%
  group_by(studyName) %>%
  dplyr::summarize(
    N_replicate = length(unique(replicate)),
    N_blockNumber = length(unique(blockNumber)),
    doRepsEqualBlocks = all(replicate == blockNumber)
  ) %>%
  ungroup() %>%
  mutate(
    CompleteBlocks = ifelse(N_replicate == 1 & N_blockNumber == 3 | N_replicate == N_blockNumber, TRUE, FALSE),
    IncompleteBlocks = ifelse(CompleteBlocks == FALSE, TRUE, FALSE)
  ) %>%
  left_join(dbdata) %>%
  mutate(
    yearInLoc = paste0(programName, "_", locationName, "_", studyYear),
    trialInLocYr = paste0(yearInLoc, "_", studyName),
    repInTrial = paste0(trialInLocYr, "_", replicate),
    blockInTrial = paste0(trialInLocYr, "_", blockNumber)
  )
```


```{r}
dbdata %>%
  dplyr::count(studyName, CompleteBlocks, IncompleteBlocks) %>%
  left_join(
    dbdata %>%
      group_by(studyName) %>%
      dplyr::summarize(
        nRepInTrial = length(unique(repInTrial)),
        nblockInTrial = length(unique(blockInTrial))
      )
  ) %>%
  gt() %>% tab_options(table.font.size = pct(75))
```

# QC Trait Values

## Density plot by trial

```{r}
dbdata %>%
  ggplot(., aes(x = FRW, fill = studyName)) +
  geom_density(alpha = 0.75) +
  guides(fill = F)

dbdata %>%
  ggplot(., aes(x = FSW, fill = studyName)) +
  geom_density(alpha = 0.75) +
  guides(fill = F)

dbdata %>%
  ggplot(., aes(x = DMCg, fill = studyName)) +
  geom_density(alpha = 0.75) +
  guides(fill = F)

dbdata %>%
  ggplot(., aes(x = DMCo, fill = studyName)) +
  geom_density(alpha = 0.75) +
  guides(fill = F)

dbdata %>%
  ggplot(., aes(x = StC, fill = studyName)) +
  geom_density(alpha = 0.75) +
  guides(fill = F)

dbdata %>%
  ggplot(., aes(x = PA, fill = studyName)) +
  geom_density(alpha = 0.75) +
  guides(fill = F)
```

## Save “cleaned” phenotypes

```{r}
saveRDS(dbdata,file=here::here("data","phenotypes_cleaned.rds"))
```
