---
title: "Phenotype Data"
author: 
  - Costa, W. G.^[Weverton Gomes da Costa, Pós-Doutorando, Embrapa Mandioca e Fruticultura, wevertonufv@gmail.com]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  out.width = "100%",
  fig.align = "center"
)
```

# Introduction

This script prepares and cleans the **phenotypic dataset** from the Embrapa Cassava Breeding Program, integrating metadata from **CassavaBase**.  
The main goals are:  
- Define the **C0 population** (founders)  
- Integrate phenotype and metadata files  
- Assign clones to populations (C0, C1, C2)  
- Check experimental design consistency (blocks, replicates, checks, tests)  
- Create nested design variables for downstream analyses  
- Perform **quality control (QC)** on trait distributions  
- Save a cleaned dataset for further modeling (BLUPs, heritability, selection indices)

---

# Step 1: Loading Packages

```{r}
library(gt)                # Elegant tables
library(tidyverse)         # Data wrangling and visualization
library(readxl)            # Import Excel files
library(data.table)        # Efficient data handling
library(genomicMateSelectR)# CassavaBase integration
```

*Comment:*  
These packages provide the backbone for data cleaning, visualization, and integration with CassavaBase formats.

---

# Step 2: Defining the C0 Population

We define the **C0 population** by excluding specific clones (checks, duplicates, uncertain IDs, blanks).  
This ensures that only true founder clones are included.

```{r}
exclud_c0 <-
  c(
    "BGM-0830-TVER",
    "BGM-0611",
    "BGM-06111",
    "BGM-0022 T.Bco",
    "BGM-0022 T.Rx",
    "BGM-0113 RcxCr",
    "BGM-0113 RcxRos",
    "BGM-0115 TB",
    "BGM-0115 TR",
    "BGM-0150A",
    "BGM-0152A",
    "BGM-0152-B",
    "BGM-0780(B)",
    "BGM-0780(E)",
    "BGM-0830",
    "BGM-0830 TB",
    "BGM-0856 CCr",
    "BGM-0856 CRx",
    "BGM-1176(M)",
    "BGM-1262C",
    "BGM-1262E",
    "BGM-1502(M)",
    "BGM-1722(D)",
    "BGM-1759(B)",
    "BGM-0150-B",
    "BGM-0082",
    "BGM-0165",
    "BGM-0177",
    "Manzhitil",
    "BGM-0271",
    "BGM-0276",
    "BGM-0312",
    "BGM-0365",
    "BGM-0890",
    "BGM-0927",
    "BGM-0931",
    "BGM-1186",
    "BGM-1212",
    "BGM-1218",
    "BGM-1490",
    "BGM-0053",
    "BGM-0307",
    "BGM-0669",
    "BGM-1376",
    "BGM-0082?",
    "BGM-0271?",
    "BGM-0276?",
    "BGM-0290?",
    "BGM-0365?",
    "BGM-0655.1?",
    "BGM-0737?",
    "BGM-0869?",
    "BGM-0895?",
    "BGM-1028?",
    "BGM-1212?",
    "BGM-0043?",
    "BGM-0144?",
    "BGM-0165.1?",
    "BGM-0165.2?",
    "BGM-0182?",
    "BGM-0187?",
    "BGM-0533?",
    "BGM-0713?",
    "BGM-0890?",
    "BGM-0931?",
    "BGM-1218?",
    "BGM-1359?",
    "Manzhitil?",
    "SE-9",
    "SE-1",
    "SE-2",
    "SE-3",
    "SE-8",
    "BGM-0655.2?",
    "BGM-0077?",
    "BGM-0929?",
    "BGM-1025?",
    "BGM-1360?",
    "BGM-1511?",
    "BLANK__G12"
  )

`%!in%` <- negate(`%in%`)

C0 <- read_delim("data/TP-BR-18-EMBRAPA-Nextgen.csv",
                 delim = ";", escape_double = FALSE, trim_ws = TRUE) %>%
  filter(tissue_id %!in% exclud_c0)
```

*Interpretation:*  
The **C0 population** represents the genetic base of the program. Removing uncertain or duplicated IDs avoids noise in downstream analyses.

---

# Step 3: Importing Phenotype and Metadata

We integrate phenotype and metadata files from CassavaBase.  
Traits are renamed to short, intuitive codes (e.g., FRW = Fresh Root Weight).

```{r}
dbdata <- readDBdata(
    phenotypeFile = here::here("data", "phenotype (1).csv"),
    metadataFile = here::here("data", "metadata (1).csv")
  ) %>%
  dplyr::rename(
    FSW = fresh.shoot.weight.measurement.in.kg.per.plot.CO_334.0000016,
    FRW = fresh.storage.root.weight.per.plot.CO_334.0000012,
    DMCg = dry.matter.content.by.specific.gravity.method.CO_334.0000160,
    DMCo = dry.matter.content.percentage.CO_334.0000092,
    NOHAV = plant.stands.harvested.counting.CO_334.0000010,
    StC = starch.content.percentage.CO_334.0000071,
    PA = plant.architecture.visual.rating.1.5.CO_334.0000099,
    DRY = dry.yield.CO_334.0000014
  ) %>%
  mutate(
    Pop = case_when(
      germplasmName %like% "BR-20GS-" | germplasmSynonyms %like% "BR-20GS-" |
      germplasmName %like% "BR-19GS-" | germplasmSynonyms %like% "BR-19GS-" ~ "C2",
      germplasmName %like% "BR-18GS-" | germplasmSynonyms %like% "BR-18GS-" ~ "C1",
      germplasmName %in% C0$tissue_id | germplasmSynonyms %in% C0$tissue_id ~ "C0"
    )
  ) %>%
  mutate_if(is.character, as.factor) %>%
  select(-numberBlocks, -numberReps)
```

*Interpretation:*  
- **C0** = founders  
- **C1** = first genomic selection cycle (2018)  
- **C2** = second genomic selection cycle (2019–2020)  

This classification allows us to track genetic gain across cycles.

---

# Step 4: Clone Counts per Population

```{r}
clone_pop <- dbdata %>% group_by(Pop, germplasmName) %>% tally()
clone_pop %>% group_by(Pop) %>% tally()
```

*Interpretation:*  
This step confirms the number of clones per population (C0, C1, C2). It ensures that population assignment was successful.

---

# Step 5: Experimental Design Checks

We count the number of blocks and replicates per trial.

```{r}
numberBlocks <- dbdata %>% group_by(studyName, blockNumber) %>% tally() %>% tally()
numberReps <- dbdata %>% group_by(studyName, replicate) %>% tally() %>% tally()
```

*Interpretation:*  
This verifies whether trials follow **complete block designs** or **incomplete block designs**.

We also remove a problematic trial:

```{r}
dbdata <- dbdata %>% filter(studyName != "BR.PTBAGRec.19.Candeal") %>% droplevels()
```

---

# Step 6: Replicates vs Blocks Consistency

We check if `replicate` corresponds to complete blocks and `blockNumber` to sub-blocks.

```{r}
dbdata %>%
  group_by(studyName) %>%
  dplyr::summarize(N_replicate = n_distinct(replicate),
                   N_blockNumber = n_distinct(blockNumber),
                   doRepsEqualBlocks = all(replicate == blockNumber))
```

*Interpretation:*  
- If `replicate == blockNumber`, the design is simple (complete blocks).  
- If not, the design includes incomplete blocks or nested structures.

---

# Step 7: Nested Design Variables

We create explicit identifiers for hierarchical design:

```{r}
dbdata %<>%
  mutate(
    yearInLoc = paste0(programName, "_", locationName, "_", studyYear),
    trialInLocYr = paste0(yearInLoc, "_", studyName),
    repInTrial = paste0(trialInLocYr, "_", replicate),
    blockInTrial = paste0(trialInLocYr, "_", blockNumber)
  )
```

*Interpretation:*  
These variables are essential for **mixed models** (BLUPs), ensuring correct nesting of random effects.

---

# Step 8: QC of Trait Values

We visualize trait distributions by trial to detect outliers or inconsistencies.

```{r}
dbdata %>% ggplot(aes(x = FRW, fill = studyName)) + geom_density(alpha = 0.75, show.legend = FALSE)
dbdata %>% ggplot(aes(x = FSW, fill = studyName)) + geom_density(alpha = 0.75, show.legend = FALSE)
dbdata %>% ggplot(aes(x = DMCg, fill = studyName)) + geom_density(alpha = 0.75, show.legend = FALSE)
dbdata %>% ggplot(aes(x = DMCo, fill = studyName)) + geom_density(alpha = 0.75, show.legend = FALSE)
dbdata %>% ggplot(aes(x = StC, fill = studyName)) + geom_density(alpha = 0.75, show.legend = FALSE)
dbdata %>% ggplot(aes(x = PA, fill = studyName)) + geom_density(alpha = 0.75, show.legend = FALSE)
```

*Interpretation:*  
Density plots allow us to quickly spot **trials with abnormal distributions**, which may indicate data entry errors or environmental effects.

---

# Step 9: Save Cleaned Dataset

```{r}
saveRDS(dbdata, file = here::here("data","phenotypes_cleaned.rds"))
```

*Interpretation:*  
The cleaned dataset is saved for downstream analyses (heritability, BLUPs, selection indices). This ensures reproducibility and avoids repeating preprocessing steps.

---

✅ **Resumo das melhorias:**  
- Comentários mais claros sobre cada etapa.  
- Interpretações biológicas/experimentais ligadas ao programa de melhoramento.  
- Estrutura narrativa que guia o leitor como um tutorial.  

---